Temporal ordering and registration of cross-sectional imaging data

### 2014-04-14 image_analysis_paper/pnas_draft.tex
Added more edits to paper
Moved things to supplement for paper

@@

\title{Unscrambling arrested development: temporal ordering and registration of cross-sectional imaging data}

@@

\begin{abstract}
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time.
%
The goal is then to temporally order the data to reconstruct the developmental dynamics.
%
If each data point is a two-dimensional image, the images must first be registered before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
%
We present an automatic approach to register {\em and} temporally order cross-sectional data sets of images.
%
The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering, and
vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
require little {\em a priori} knowledge of the image features or the developmental dynamics.
%
We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
\end{abstract}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained for Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in (b).}
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

%%%
%%% orientation does not read good to Y below -
%%% it is the same orientation, but a different rotation around the axis
%%% how to say this ?
%%%
Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
%
The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
%
Both of these tasks can be fully automated, and even combined.
%
We will use {\em angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use {\em diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using {\em vector diffusion maps} \cite{singer2012vector}.
%
%The paper is organized as follows.
%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
%
%Then, we show that both of these tasks can be effectively combined in a single step.
%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
%
%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
%
%However, this algorithm requires distances between data points.
%
%Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
%
%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.

@@

We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
%
{\bf It is not clear how we are rotating the arrows (around the center vs. around the tail).}
%
From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
provides the rotation angles (up to a constant rotation) assigned to each data point (see {\it Materials and Methods}).
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The pariwise alignment angles are indicated. (b) The vectors from (a), each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. (d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:synch1}{\ref{fig:schematics}a}
\customlabel{subfig:synch2}{\ref{fig:schematics}b}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
\end{figure}
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the peaks appear aligned,
even though the algorithm uses no specific information about the existence or location of such peaks.
%
%We note once more that, since angular synchronization only uses pairwise rotation information,
%the optimal rotations are recovered up to a global rotation.

@@

After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
%
%Our main assumption is that most of the variation in the data correlates with time.
%
Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
%
Figure~\ref{subfig:dmaps_edges} shows a schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
We assume that the data points are ordered from youngest to oldest along the curve.
%
When our data are very high dimensional, we cannot extract this curve by eye.
%
We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading (non-trivial) diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.

To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored based on the distance between the data points (using the diffusion kernel of the distance), so that pairs of points which are close have dark/strong edges connecting them.
%
Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
%
%This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

@@

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the registrations {\em and} the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion map coordinate.

@@

The same methods (with small modifications) can be applied directly to our two-dimensional images.
%
Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
%
To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\em three-dimensional rotations} (symmetry group $SO(3)$; see {\it Materials and Methods}, {\it SI Appendix}).
%
We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
the results are shown in Figure~\ref{fig:images_ordered}.
%
Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.

@@

We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
%
Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
%
To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to this extent.
%
Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.

Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
%
Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
%
Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
here, each data point is a vector containing the expression levels of the various genes.
%
Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
%
These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
%
The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
expecting that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
%
We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
%
Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
%
They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
%
{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
%%%
%%%   we need help here
%%%
%
%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
%
%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.

The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
%
For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
%
Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
%
In general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
%
Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
%
However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
%
{\bf Need to discuss consistency.}

Angular synchronization and vector diffusion maps require no feature points and no template function.
%
These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
%
In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics. % (which underpins the development of vector diffusion map algorithms).
%
In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
%
Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
%
Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.

@@

Let $x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
each signal is a function defined on the unit circle (on the plane).
%
%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
First assume that each signal $x_i$ is a {\em noisy} rotated copy of the underlying signal $x_{true}$
(which we are {\em not} given), such that
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
%
Our goal is to recover $\theta_1, \dots, \theta_m$.
%
Up to noise,
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
For the one-dimensional discretized profiles shown in Figure~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
%
Alternatively, for continuous signals, an optimization algorithm
can be used \cite{ahuja2007template}.

@@

Using angular synchronization or vector diffusion maps to register images with respect to rotations {\em and translations} requires some additional effort.

@@

%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations.

>>>>

%\DeclareMathSizes{9}{8}{7}{7}

@@

\title{Temporal ordering and registration of cross-sectional imaging data}

@@

\begin{abstract}
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time.
%
The goal is then to temporally order the data to reconstruct the developmental dynamics.
%
If each data point is a two-dimensional image, the images must first be registered before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
%
We present an automatic approach to register {\it and} temporally order cross-sectional data sets of images.
%
The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering, and
vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
require little {\it a priori} knowledge of the image features or the developmental dynamics.
%
We demonstrate the utility of our methods using a collection of images from a study of {\it Drosophila} embryogenesis.
\end{abstract}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\it Drosophila} embryos. Each embryo has been stained for Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in (b).}
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

%%%
%%% orientation does not read good to Y below -
%%% it is the same orientation, but a different rotation around the axis
%%% how to say this ?
%%%
Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
%
The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
%
Both of these tasks can be fully automated, and even combined.
%
We will use {\it angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use {\it diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using {\it vector diffusion maps} \cite{singer2012vector}.
%
%The paper is organized as follows.
%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
%
%Then, we show that both of these tasks can be effectively combined in a single step.
%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
%
%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
%
%However, this algorithm requires distances between data points.
%
%Because each image is in a different orientation, we need to first align or {\it register} the images before computing distances.
%
%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.

@@

We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\it a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
%
{\bf It is not clear how we are rotating the arrows (around the center vs. around the tail).}
%
From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\it most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
provides the rotation angles (up to a constant rotation) assigned to each data point (see {\it Materials and Methods}).
%
%Because the method considers {\it all} pairwise alignments, angular synchronization is often robust to noise.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The pariwise alignment angles are indicated. (b) The vectors from (a), each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. (d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:synch1}{\ref{fig:schematics}a}
\customlabel{subfig:synch2}{\ref{fig:schematics}b}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
\end{figure}
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the peaks appear aligned,
even though the algorithm uses no specific information about the existence or location of such peaks.
%
%We note once more that, since angular synchronization only uses pairwise rotation information,
%the optimal rotations are recovered up to a global rotation.

@@

After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
%
%Our main assumption is that most of the variation in the data correlates with time.
%
Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\it and} that this curve is parameterized by time.
%
Figure~\ref{subfig:dmaps_edges} shows a schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
We assume that the data points are ordered from youngest to oldest along the curve.
%
When our data are very high dimensional, we cannot extract this curve by eye.
%
We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading (non-trivial) diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.

To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\it on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\it as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored based on the distance between the data points (using the diffusion kernel of the distance), so that pairs of points which are close have dark/strong edges connecting them.
%
Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
%
%This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

@@

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\it simultaneously} recover the registrations {\it and} the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion map coordinate.

@@

The same methods (with small modifications) can be applied directly to our two-dimensional images.
%
Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
%
To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\it three-dimensional rotations} (symmetry group $SO(3)$; see {\it SI Appendix}).
%
We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
the results are shown in Figure~\ref{fig:images_ordered}.
%
Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.

@@

We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
%
Although we illustrated our methods using a specific data set from a study of {\it Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
%
To our knowledge, temporal ordering of {\it Drosophila} images has not been automated to this extent.
%
Recently, machine learning algorithms have been used to group {\it Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.

Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
%
Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
%
Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
here, each data point is a vector containing the expression levels of the various genes.
%
Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
%
These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
%
The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
%
We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
%
Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
%
They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
%
{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
%%%
%%%   we need help here
%%%
%
%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
%
%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.

The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
%
For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
%
Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
%
In general, appropriate features may not be known {\it a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
%
Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
%
However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
%
{\bf Need to discuss consistency.}

Angular synchronization and vector diffusion maps require no feature points and no template function.
%
These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\it same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
%
In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics. % (which underpins the development of vector diffusion map algorithms).
%
In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
%
Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
%
Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.

@@

Let $x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
each signal is a function defined on the unit circle (on the plane).
%
%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
First assume that each signal $x_i$ is a {\it noisy} rotated copy of the underlying signal $x_{true}$
(which we are {\it not} given), such that
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
%
Our goal is to recover $\theta_1, \dots, \theta_m$.
%
Up to noise,
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\it estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
For the one-dimensional discretized profiles shown in Figure~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
%
Alternatively, for continuous signals, an optimization algorithm
can be used \cite{ahuja2007template}.

@@

Using angular synchronization or vector diffusion maps to register images with respect to rotations {\it and translations} requires some additional effort.

@@

%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\it and} translations.
